---
title: "Characterisation of a Putative Agent of Horizontal Gene Transfer Encoded by a Bacterial Prophage Region"
author: "Y3877831"
output:
    html_document:
    css: css/bootstrap-minty.css
    toc: true
    depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
    number_sections: true
    highlight: null
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    include = FALSE,
    echo = FALSE,
    fig.align = "center",
    fig.retina = 3,
    message = FALSE,
    warning = FALSE
)
```

```{r pkgs}
source("../R/00-pkgs.R", local = knitr::knit_global())
```

```{r write-pkgs-bib}
knitr::write_bib(
    c(
        "base",
        "RefManageR",
        "tidyverse",
        "rmarkdown",
        "plotly",
        "ggpubr",
        "htmltools"
    ),
    width = 80,
    prefix = "",
    file = "../bibliography/pkgs.bib"
)
```

```{r import-bibliography}
RefManageR::BibOptions(
    check.entries = FALSE,
    bib.style = "authoryear",
    cite.style = "authoryear",
    style = "markdown",
    hyperlink = FALSE,
    dashed = FALSE,
    longnamesfirst = FALSE,
    max.names = 3
)

refs <-
    RefManageR::ReadBib("../bibliography/refs.bib", check = FALSE)
pkgs <-
    RefManageR::ReadBib("../bibliography/pkgs.bib", check = FALSE)
```


# Introduction
MuGam is a homologue of eukaryotic proteins Ku70/Ku80, which are involved in recognizing double-strand DNA prior to repair, as well as DNA transposition. In *Haemophilus influenzae*, Gam (HiGam) inhibits digestion of Mu phage DNA by nucleases `r Cite(refs, "DAdda_di_Fagagna2003-kb")`.

The aim of this report is to characterise HiGam by processing the data obtained from the following experiments:

* Polymerase chain reaction (PCR)
* Quantitative polymerase chain reaction (qPCR)
* Electrophoretic mobility shift assay (EMSA)
* Size-exclusion chromatography (SEC)
* Analytical ultracentrifugation (AUC) ???


___
# Requirements
```{r ggplot-themes}
source("../R/themes.R", local = knitr::knit_global())
```

```{r functions}
source("../functions/count_words.R", local = knitr::knit_global())
source("../functions/relative_front.R", local = knitr::knit_global())
source("../functions/pred_intercept_sec.R", local = knitr::knit_global())
source("../functions/pred_intercept_emsa.R", local = knitr::knit_global())
source("../functions/nls.R", local = knitr::knit_global())
```

```{r data-import}
# Load the PCR data
pcr <- read.csv("../data/data-raw/pcr.csv", header = TRUE) %>%
  janitor::clean_names()

# Load the PCR calibration data for the gDNA library
gdna_calibration <-
  read.csv("../data/data-raw/pcr_gdna_calibration.csv", header = TRUE) %>%
  janitor::clean_names()

# Load the PCR calibration data for the cDNA library
cdna_calibration <-
  read.csv("../data/data-raw/pcr_cdna_calibration.csv", header = TRUE) %>%
  janitor::clean_names()

# Load the EMSA data
emsa <- read.csv("../data/data-raw/emsa.csv", header = TRUE) %>%
  janitor::clean_names()

# Load the EMSA calibration data
emsa_calibration <-
  read.csv("../data/data-raw/emsa_calibration.csv", header = TRUE) %>%
  janitor::clean_names()

# Load the SEC standards data
standards <-
  read.csv("../data/data-raw/sec_std.csv", header = TRUE) %>%
  janitor::clean_names()

# Load the SEC data
sec <-
  read.csv("../data/data-raw/sec_im9higam.csv", header = TRUE) %>%
  janitor::clean_names()

```

```{r data-tidying}
standards <- standards %>%
    rename(
        c(
            "Ferritin" = curve_ferritin,
            "IgG" = curve_ig_g,
            "Myoglobin" = curve_myoglobin,
            "Ovalbumin" = curve_ovalbumin,
            "Thyroglobulin" = curve_thyroglobulin,
            "Transferrin" = curve_transferrin,
            "Vitamin B12" = curve_vitamin_b12
        )
    ) %>%
    pivot_longer(cols = !c(x, vol),
                 names_to = "protein",
                 values_to = "absorbance_units")

sec <- sec %>%
    pivot_longer(cols = starts_with("a"),
                 names_to = "protein",
                 values_to = "absorbance")

# Separate the PCR data set by DNA library for analysis
gdna <- pcr %>% # gDNA PCR data
    filter(library == "gDNA")

cdna <- pcr %>% # cDNA
    filter(library == "cDNA")
```

___
# Analysis 

## Linear Regression for gDNA and cDNA PCR

```{r pcr-linear-models}
#gDNA
gdna_mod <- lm(data = gdna_calibration, log ~ rf)

#cDNA
cdna_mod <- lm(data = cdna_calibration, log ~ rf)
```

```{r pcr-preperation}
# gDNA
gdna <- gdna %>%
  mutate(
    rf = relative_front(gdna),
    log = ((rf * gdna_mod$coefficients[2]) + gdna_mod$coefficients[1]),
    actual_size = (10 ^ log),
    diff = ((expected_size - actual_size) / expected_size * 100)
  )

# cDNA
cdna <- cdna %>%
  mutate(
    rf = relative_front(cdna),
    log = ((rf * cdna_mod$coefficients[2]) + cdna_mod$coefficients[1]),
    actual_size = (10 ^ log),
    diff = ((expected_size - actual_size) / expected_size * 100)
  )

# join gDNA & cDNA
pcr_full <- full_join(gdna, cdna, by = NULL)
```

```{r pcr-plot, include = TRUE}
pcr_plot <- pcr_full %>%
    ggplot(aes(
        x = gene,
        y = diff,
        fill = library,
        text = paste(
            "Library: ",
            library,
            "<br>Gene: ",
            gene,
            "<br>Difference: ",
            round(diff),
            "%",
            "<br>Expected Size (bp): ",
            expected_size,
            "<br>Actual Size (bp):",
            round(actual_size)
        )
    )) + geom_col(position = "dodge") +
    geom_hline(yintercept = 0, size = 0.25) +
    mytheme() +
    scale_fill_brewer(name = "DNA Library",
                      labels = c("cDNA", "gDNA"),
                      palette = "Set3",
    ) +
    labs(y = "Percentage Difference to Expected Size", x = "Gene")


htmltools::div(ggplotly(pcr_plot, tooltip = c("text")), align = "center")
```


## Electrophoretic Mobility Shift Assay (EMSA) & Binding Affinity
```{r emsa-linear-model}
emsa_mod_plot <-
    lm(data = emsa_calibration, log ~ rf) %>% plot(which = 1)
```

```{r emsa-calibrarion-plot}
emsa_calibration_plot <- emsa_calibration %>%
    ggplot(aes(x = rf, y = log)) +
    geom_point(shape = 18,
               size = 2,
               colour = "#D23681") +
    ggtitle("Calibration Plot for EMSA Migration") +
    scale_x_continuous(
        expand = c(0.02, 0.02),
        breaks = seq(0, 1, by = 0.05),
        name = expression(paste("R"["f"]))
    ) +
    scale_y_continuous(
        expand = c(0, 0.2),
        breaks = seq(5.4, 7, by = 0.2),
        name = "log(DNA Molecular Weight in bp)"
    ) +
    geom_smooth(
        method = lm,
        formula = y ~ x,
        se = FALSE,
        colour = "#29a198",
        linetype = 2,
        size = 0.5
    ) +
    mytheme()
```

```{r nls}
emsa_fit <- NLS(emsa)
```

```{r extract-pred}
rf_max = summary(emsa_fit)$coefficients[1, 1]
kd = summary(emsa_fit)$coefficients[2, 1]
```

```{r emsa-diff}
# Calculate the difference between the measured rf and predicted rf value
# This is needed for the residuals
emsa_pred <- predict(emsa_fit)

# Calculate residuals to the fit
emsa_res <- emsa %>%
    mutate(residuals = rf - emsa_pred)
```

```{r emsa-res-plot, include = TRUE}
emsa_plot <- emsa %>%
  ggplot(aes(x = conc, y = rf)) +
  geom_point(shape = 20,
             size = 3,
             colour = "black") +
  stat_function(fun = pred_intercept_emsa, colour = "#66C2A4") +
  mytheme() +
  scale_y_continuous(limits = c(0, 0.6),
                     breaks = seq(0, 7, by = 0.10)) +
  scale_x_continuous(
    expand = c(0.02, 0.02),
    limits = c(0, 8),
    breaks = seq(0, 8, by = 1)
  ) +
  xlab(NULL) +
  ylab(expression(paste(bold(
    "Relative Front (a.U.)"
  ))))

# Plot the residuals and a line showing y = 0
emsa_res_plot <- emsa_res %>%
  ggplot(aes(x = conc, y = residuals)) +
  geom_point(shape = 20,
             size = 3,
             colour = "black") +
  geom_hline(aes(yintercept = 0),
             col = '#66C2A4',
             size = .5) +
  scale_x_continuous(
    expand = c(0.02, 0.02),
    limits = c(0, 8),
    breaks = seq(0, 8, by = 1),
    name = (expression(bold(
      atop("Protein Concentration (μM)", "Im9"["dimer"])
    )))
  ) +
  scale_y_continuous(
    expand = c(0.02, 0.02),
    limits = c(-0.1, 0.1),
    breaks = seq(-0.1, 0.1, by = 0.1),
    name = "Residual (a.U.)"
  ) + mytheme()


emsa_grid_plot <- ggarrange(emsa_plot, emsa_res_plot, nrow = 2, ncol = 1, widths = c(1, 1/2))

emsa_grid_plot

htmltools::div(emsa_grid_plot, align = "center")
```



## Size Exclusion Chromatography (SEC)
```{r standards-plot, include = TRUE}
standards_plot <- standards %>%
    ggplot(aes(
        x = vol,
        y = absorbance_units,
        colour = protein),
    show.legend = TRUE) +
    geom_line() +
    scale_colour_brewer(name = "Protein",
                        palette = "Set2") + # colour blind friendly
    mytheme() +
    scale_x_continuous(
        expand = c(0, 0),
        limits = c(9, 22),
        breaks = seq(9, 22, by = 1),
        name = "Elution Volume (mL)"
    ) + scale_y_continuous(
        expand = c(0, 0),
        limits = c(0, 100),
        breaks = seq(0, 100, by = 10),
        name = "A280 Absorbance Units (A.U.)"
    )


htmltools::div(ggplotly(standards_plot, align = "center"))
```

```{r mw-ev}
mw_ev <- standards %>%
    group_by(protein) %>%
    filter(absorbance_units == max(absorbance_units)) %>%
    summarise(vol) %>%
    ungroup() %>%
    mutate(
        mw = recode(
            protein,
            "Thyroglobulin" = 669000,
            "Ferritin" = 440000,
            "IgG" = 150000,
            "Transferrin" = 81000,
            "Ovalbumin" = 43000,
            "Myoglobin" = 17000,
            "Vitamin B12" = 1355
        ),
        logmw = log(mw)
    )

mw_ev
```

```{r sec-lm}
linear_reg <- lm(mw_ev$vol ~ mw_ev$logmw)
intercept <- linear_reg$coefficients[1] # intercept value
slope <- linear_reg$coefficients[2] # slope value
```

```{r calibration-plot, include = TRUE}
calibration_plot <- mw_ev %>%
    ggplot(mapping = aes(y = vol, x = logmw)) +
    geom_point() +
    mytheme() +
    stat_function(fun = pred_intercept_sec,
                  colour = "#8D9FCA") +
    theme(plot.margin = unit(c(1, 1, 1, 1), "cm")) +
    scale_x_continuous(
        expand = c(0.05, 0.05),
        limits = c(7, 14),
        breaks = seq(7, 14, by = 1),
        name = "log (MW)"
    ) + scale_y_continuous(
        expand = c(0.05, 0.05),
        limits = c(9, 22),
        breaks = seq(9, 22, by = 2),
        name = "Elution Volume (mL)"
    )

calibration_plot
```

```{r sec-plot, include = TRUE}
sec_plot <- sec %>%
    ggplot(aes(
        x = vol,
        y = absorbance,
        colour = protein),
    show.legend = TRUE) +
    geom_line() +
    scale_colour_brewer(name = "Protein",
                        labels = c("a280", "a254"),
                        palette = "Set2") +
    scale_x_continuous(limits = c(13, 17)) +
    scale_y_continuous(expand = c(0, 0),
                       limits = c(0, 100)) +
    mytheme() +
    xlab("Elution Volume (mL)") +
    ylab("Absorbance (mAU)")


htmltools::div(ggplotly(sec_plot, align = "center"))
```

```{r sec-mw}
# Getting the elution volume that corresponds to the maximum value of a curve

volgam <- sec %>%
    group_by(protein) %>%
    filter(vol == max(vol))

# Estimated MW from the 280 nm data
a280_mw <- exp((volgam$time[1] - intercept) / slope)

# Estimated MW from the 254 nm data
a254_mw <- exp((volgam$time[2] - intercept) / slope)
```

# {-}
___
**This Document:** `r count_words("report.Rmd")`\
**README:** `r count_words("../README.md")`\
**Total:** `r count_words("report.Rmd") + count_words("../README.md")`\

___
Report made with `rmarkdown` `r Cite(pkgs, c("rmarkdown","rmarkdown2018", "rmarkdown2020"))`, `RefManageR``r Cite(pkgs, "RefManageR")`.\

Theme is *Minty* by [Bootswatch](https://bootswatch.com/minty/).

___
# Bibliography {.tabset}

## Packages {-}
```{r pkgs-bib, include = TRUE, results = "asis"}
RefManageR::PrintBibliography(pkgs)
```

## References {-}
```{r refs-bib, include = TRUE, results =  "asis"}
RefManageR::PrintBibliography(refs)
```
